---
title: "Analysis on Transmission Type"
author: "Justin Pizzino"
date: "8/15/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stats)
library(ggplot2)
library(car)
library(corrplot)
```

## R Markdown
First load the data, then convert the incorrect data to factors, such as am

```{r data}
data(mtcars)
#which items should be factors?  
mtcars<-mtcars
mtcars$am<-as.logical(mtcars$am)
#V2mtcars$cyl<-as.factor(V2mtcars$cyl)
#mtcars$cyl.4<-as.logical(FALSE)  #There is a better way to do this
#mtcars$cyl.4[cyl=4]<-as.logical(TRUE)
mtcars$cyl.6<-as.logical(FALSE)  #There is a better way to do this
mtcars$cyl.6[ which(mtcars$cyl==6)]<-as.logical(TRUE)
mtcars$cyl.8<-as.logical(FALSE)  #There is a better way to do this
mtcars$cyl.8[ which(mtcars$cyl==8)]<-as.logical(TRUE)
```


Easy way to start is to do a direct compare of mpg to the transmission type.  From here it woudl appear that a manual tranmisssion would add about 7.2 MPG.  But we know there is more to it than that.
```{r start}
summary(lm(mpg~am,mtcars))$coef
```


We know we should include other data, but need to determine which data points.  So I compraed the corrlation of each data point to it self.
```{r cor}
corrplot(cor(mtcars),method="square")
```